Tutorial: Quantifying Spatially Resolved Hydration Thermodynamics Using Grid Inhomogeneous Solvation Theory [Article v\versionnumber]

### 2024-07-11 manuscript/manuscript.tex

\corr{klaus.liedl@uibk.ac.at}{KRL}
\corr{mgilson@health.ucsd.edu}{MKG}
\corr{thomas.kurtzman@lehman.cuny.edu}{TK}
%TODO Vjays ORCID is missing
\orcid{Valentin J Hoerschinger}{0000-0002-4469-3238}
\orcid{Franz Waibl}{0000-0003-0527-0803}
\orcid{Vjay Molino}{}
\orcid{Helmut Carter}{0000-0003-0273-4107}
\orcid{Monica L Fern{\'a}ndez-Quintero}{0000-0002-6811-6283}
\orcid{Steven Ramsey}{0000-0001-7441-3228}
\orcid{Daniel R Roe}{0000-0002-5834-2447}
\orcid{Klaus R Liedl}{0000-0002-0985-2299}
\orcid{Michael K Gilson}{0000-0002-3375-1738}
\orcid{Tom Kurtzman}{0000-0003-0900-772X}

@@

In this tutorial, we aim to guide the reader through the most common steps involved in a GIST analysis using the streptavidin-biotin complex as an demonstrative system. Towards this, Jupyter notebooks and Python packages are provided to simplify the analysis.
Furthermore, we discuss the theory of GIST with a focus on practical aspects.
We highlight potential pitfalls and show how to avoid technical difficulties.
For this tutorial, familiarity with molecular dynamics simulations as well as the AmberTools package \cite{Case2023-ambertools} is assumed.

%This particular document provides a skeleton illustrating key sections for a Tutorial document.
%Please see the sample \texttt{sample-document.tex} in \url{github.com/livecomsjournal/article_templates/templates} for additional information on and examples of using the LiveCoMS LaTeX class.
%Here we also assume familiarity with LaTeX and knowledge of how to include figures, tables, etc.; if you want examples, see the sample just referenced.
%
%In your work, in this particular slot, please provide an abstract of no more than 250 words.
%Your abstract should explain the main contributions of your article, and should not contain any material that is not included in the main text.
%Please note that your abstract, plus the authorship material following it, must not extend beyond the title page or modifications to the LaTeX class will likely be needed.
\end{abstract}

@@

%Tutorials should endeavor to cover the specific task at hand, and also highlight how the steps might need to be modified (or additional care might need to be taken at particular points) to handle more general cases.
%
%The scope of the tutorial, as well as the expected proficiencies / outcomes for researchers who complete the tutorial, should be clearly defined.
%This will often happen in a specific section or subsection in the article itself.

@@

\pagebreak %TODO: maybe remove this by better formatting

@@

\subsubsection{Further steps}
In this tutorial, we obtained a value for the binding of biotin into the closed conformation of the streptavidin binding pocket.
To investigate the effect of different conformations on the binding affinity, you can perform molecular dynamics simulations of the complex and/or monomers and perform GIST on multiple cluster representatives.
To include the effect of lid closing in the binding process, free energy calculation methods such as umbrella sampling can be used.
To speed up the calculation, smaller GIST grids could be used by focusing on the binding pocket, or by rotating the solute molecule by its principal axes to fit the cuboid grid more exactly.
However, this should be done \emph{before} the MD simulation, since rotating the trajectory damages the periodic box information.
%TODO: Ask Dan if this is still the case

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{streptavidin_bp_close.png}
	\caption{Volume of streptavidin defined around \SI{3.5}{\angstrom} of biotin. }\label{streptavidin_volume}
\end{figure}
%TODO: Hide capping group
\pagebreak %TODO pagebreak
\section{Theory}
\label{sec:theory}
GIST is an implementation of inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998} that discretizes the free energy of solvation $\dgsolv$ on a three-dimensional grid. 
It was first devised by Nguyen et al.\ \cite{Nguyen2012} and its implementation in \software{cpptraj} was thoroughly described by Ramsey et al.\ \cite{Ramsey2016}.
Here, we only provide a short overview of the theory behind GIST.
For more technical information (such as implementation details), we recommend one of the more recent publications on developments in GIST \cite{Kraml2020,Chen2021,Roe2023-mpi-gist}.

@@

\begin{equation}
	\Delta E^\text{R}_{ww} = \sum_{k}^{\text{voxels in }R} E_{ww,k}^\text{corr}
%	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
\end{equation}
When the region $R$ comprises more than one solvent molecule, interactions within this region are double-counted while interactions to the outside are not.
This could be solved using the solvent-solvent energy between each pair of voxels $E_{ww,k\;l}$:
\begin{equation}
	\Delta E^\text{R}_{ww} = 2\left(\sum_{k}^\text{voxels in R} E_{ww,k}^\text{corr}-\sum_{k}^\text{voxels in R}\sum_{l}^\text{l>k} E_{ww, k\;l}\right)
\end{equation}
While this is supported by the standard GIST implementation (using the \inlinecode{doeij} flag), it is rarely done due to the large size of the $E_{ww,k\;l}$ matrix.
Integrating GIST values over the whole grid corresponds to a process where all the solvent is removed into bulk.
However, when integrating over a small region, this process is less well-defined, since it is unclear to what extent the remaining solvent would reorganize.
This depends on the local environment and on \emph{what} the solvent would be replaced by.
Therefore, the effect of reorganization must be judged for each case individually.
One way to treat reorganization is an end-state approach as shown in the tutorial: all states before and after solvent reorganization are considered. In the tutorial, this corresponds to the solvation of biotin and streptavidin seperated (initial states) and in complex (end-state). Note that all solvent contributions need to be considered to get thermodynamically accurate results in such a case. In practice this therefore requires the integration over the whole grid or at least all voxels with solvent properties different from bulk (i.e. setting a large cutoff).
\subsubsection{PME energy}
In the original version of GIST, the energies are calculated based on the the minimum image convention.
In PME-GIST, the electrostatic energy $E_\text{elec}$ is calculated using the particle mesh Ewald (PME) method, which yields energies that are highly consistent with the Amber MD engine \cite{Chen2021}.
The Lennard-Jones part $E_\text{lj}$ is computed separately in direct space.

>>>>

\corr{klaus.liedl@uibk.ac.at}{KRL}
\corr{mgilson@health.ucsd.edu}{MKG}
\corr{thomas.kurtzman@lehman.cuny.edu}{TK}
\orcid{Valentin J Hoerschinger}{0000-0002-4469-3238}
\orcid{Franz Waibl}{0000-0003-0527-0803}
\orcid{Vjay Molino}{0009-0005-2055-9295}
\orcid{Helmut Carter}{0000-0003-0273-4107}
\orcid{Monica L Fern{\'a}ndez-Quintero}{0000-0002-6811-6283}
\orcid{Steven Ramsey}{0000-0001-7441-3228}
\orcid{Daniel R Roe}{0000-0002-5834-2447}
\orcid{Klaus R Liedl}{0000-0002-0985-2299}
\orcid{Michael K Gilson}{0000-0002-3375-1738}
\orcid{Tom Kurtzman}{0000-0003-0900-772X}

@@

In this tutorial, we aim to guide the reader through the most common steps involved in a GIST analysis using the streptavidin-biotin complex as a demonstrative system. Towards this, Jupyter notebooks and Python packages are provided to simplify the analysis.
Furthermore, we discuss the theory of GIST with a focus on practical aspects.
We highlight potential pitfalls and provide strategies to avoid technical difficulties.
This tutorial assumes familiarity with molecular dynamics simulations and the AmberTools package.

\end{abstract}

@@

@@

\pagebreak %TODO: pagebreak, maybe remove these by better formatting

@@

\subsubsection{Further steps}
In this tutorial, we obtained a value for the binding of biotin into the closed conformation of the streptavidin binding pocket.
To investigate the effect of different conformations on the binding affinity, you can perform molecular dynamics simulations of the complex and/or monomers and perform GIST on multiple cluster representatives.
To include the effect of lid closing in the binding process, free energy calculation methods such as umbrella sampling can be used.
To speed up the calculation, smaller GIST grids could be used by focusing on the binding pocket, or by rotating the solute molecule by its principal axes to fit the cuboid grid more exactly.
However, this should be done \emph{before} the MD simulation, since rotating the trajectory damages the periodic box information.

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{streptavidin_bp_close.png}
	\caption{Volume of streptavidin defined around \SI{3.5}{\angstrom} of biotin. }\label{streptavidin_volume}
\end{figure}

\pagebreak %TODO pagebreak
\section{Theory}
\label{sec:theory}
GIST is an implementation of inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998} that discretizes the free energy of solvation $\dgsolv$ on a three-dimensional grid. 
It was first devised by Nguyen et al.\ \cite{Nguyen2012} and its implementation in \software{cpptraj} was thoroughly described by Ramsey et al.\ \cite{Ramsey2016}.
Here, we only provide a short overview of the theory behind GIST.
For more technical information (such as implementation details), we recommend one of the more recent publications on developments in GIST \cite{Kraml2020,Chen2021,Roe2023-mpi-gist}.

@@

\begin{equation}
	\Delta E^\text{R}_{ww} = \sum_{k}^{\text{voxels in }R} E_{ww,k}^\text{corr}
%	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
\end{equation}
When the region $R$ comprises more than one solvent molecule, interactions within this region are double-counted while interactions to the outside are not.
This could be solved using the solvent-solvent energy between each pair of voxels $E_{ww,k\;l}$:
\begin{equation}
	\Delta E^\text{R}_{ww} = 2\left(\sum_{k}^\text{voxels in R} E_{ww,k}^\text{corr}-\sum_{k}^\text{voxels in R}\sum_{l}^\text{l>k} E_{ww, k\;l}\right)
\end{equation}
While this is supported by the standard GIST implementation (using the \inlinecode{doeij} flag), it is rarely done due to the large size of the $E_{ww,k\;l}$ matrix.
Integrating GIST values over the whole grid corresponds to a process where all the solvent is removed into bulk.
However, when integrating over a small region, this process is less well-defined, since it is unclear to what extent the remaining solvent would reorganize.
This depends on the local environment and on \emph{what} the solvent would be replaced by.
Therefore, the effect of reorganization must be judged for each case individually.
One way to treat reorganization is an end-state approach as shown in the tutorial: all states before and after solvent reorganization are considered. In the tutorial, this corresponds to the solvation of biotin and streptavidin seperated (initial states) and in complex (end-state). Note that all solvent contributions need to be considered to get thermodynamically accurate results in such a case. In practice this therefore requires the integration over the whole grid or at least all voxels with solvent properties different from bulk (i.e. setting a large cutoff).
\subsubsection{PME energy}
In the original version of GIST, the energies are calculated based on the minimum image convention.
In PME-GIST, the electrostatic energy $E_\text{elec}$ is calculated using the particle mesh Ewald (PME) method, which yields energies that are highly consistent with the Amber MD engine \cite{Chen2021}.
The Lennard-Jones part $E_\text{lj}$ is computed separately in direct space.
